Quantifying and controlling biases in dark matter halo concentration estimates

### 2016-09-01 paper/ms.tex
shorter caption
updated draft

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods on the Bolshoi data using only relaxed halos.
  The lines correspond to the median concentration values in each mass
  bin.  
  The shaded region presents 10 and 90 per cent spread. 
  The three methods have a similar spread but for clarity we only show
  the spread for the new method.   
  The dotted line corresponds to fits reported by \citep{Prada2012}.
  The left panel shows the raw results coming from each algorithm. 
  The right panel introduces a correction on the velocity and
  integrated mass results following the results of the downsampling
  experiments presented in Figure 
  \ref{fig:downsampling}.
  \label{fig:concentration}} 
\end{figure*}

@@

We take halos with at least $10^{5}$ particles and subsample them by
factors of $2$ up to $10^{3}$. 
We measure the concentration at every resampling.
We use a two-sample Kolmogorov-Smirnov (KS) test to  compare the list
of radial distances from each subsample against that of its parent halo.
We find that the resulting p-value distribution is flat.
This confirms that the radial particle distribution in the bootraped
halo is consistent with coming from the distribution given by the
parent halo.

@@

Figure \ref{fig:downsampling} summarizes our results.  
The plot on the left shows the average value of $f_{\rm off}$ as a
function of particle number.  
This can be interpreted as the statistical bias on the concentration
estimate. 
For large enough particle numbers, $N_{s}>4\times 10^3$ the results of
the three algorithms show a bias below the $1\%$ level.
For a lower number of particles the results start to deviate.
At $200$ particles the velocity method overestimates the concentration
by a factor $14\%$ while the density method overestimates it by $20\%$.
Around the same sampling scale, the new algorithm shows a more stable
behaviour underestimating the concentration only by a factor of
$1\%$-$3\%$.
Work  that studied the effect of
lowering numerical resolution on the same individual halos
\citep{2008MNRAS.391.1685S} yielded similar values for the
concentration across resolutions.  That work  explored a particle
number range $>5\times 10^5$, completely consistent with our result that the
bias can only be evident for particle numbers $<4\times 10^3$.

>>>>

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods on the Bolshoi data using only relaxed halos.
  The lines correspond to the median concentration values in each mass
  bin.  
  The shaded region presents 10 and 90 per cent spread. 
  The three methods have a similar spread but for clarity we only show
  the spread for the new method.   
  The dotted line corresponds to fits reported by \citep{Prada2012}.
  The left panel shows the raw results coming from each algorithm. 
  The right panel introduces a correction following the results on the
  bias as a function of particle number.
  \label{fig:concentration}} 
\end{figure*}

@@

We take halos with at least $10^{5}$ particles and subsample them by
factors of $2$ up to $10^{3}$. 
We measure the concentration at every resampling.
We use a two-sample Kolmogorov-Smirnov (KS) test to  compare the list
of radial distances from each subsample against that of its parent halo.
We find that the resulting p-value distribution is flat.
This confirms that the radial particle distribution in the bootraped
halo is consistent with coming from the distribution given by the
parent halo.  
Why not using different simulations with the same initial conditions
and lower resolutions \citep[i.e.][]{2008MNRAS.391.1685S}?. 
Because we want to be sure that we are only measuring the bias of a
given method as a function of particle number for statistically
identical halos, and not a possible simulation artifact that changes
the halo structure.

@@

Figure \ref{fig:downsampling} summarizes our results.  
The plot on the left shows the average value of $f_{\rm off}$ as a
function of particle number.  
This can be interpreted as the statistical bias on the concentration
estimate. 
For large enough particle numbers, $N_{s}>4\times 10^3$ the results of
the three algorithms show a bias below the $1\%$ level.
For a lower number of particles the results start to deviate.
At $200$ particles the velocity method overestimates the concentration
by a factor $14\%$ while the density method overestimates it by $20\%$.
Around the same sampling scale, the new algorithm shows a more stable
behaviour underestimating the concentration only by a factor of
$1\%$-$3\%$.
